---
title: "lung-cancer-outlier-removal"
author: "Hege MB"
date: "11 Jan 2016"
output: html_document
---

Gene expression data from the lung cancer study was made available on Stallo January 11 2016. The data is updated from the Cancer Registry through Dec 31 2011.

```{r import libraries}
#############################################
## import libaries
#############################################
require(lumi)
require("arrayQualityMetrics")
options(bitmapType="cairo")
require(grid) #for plots with raster
require(ggplot2)
require("gridExtra")
require(pandoc)
options(bitmapType='cairo')
```


```{r load data}
#setwd("/home/hbo010/lung_cancer/scripts")
resultPath <- "/home/hbo010/lung_cancer/results"
## Load the R-object:
load("/project/data1/tice/GCF-2015-373_LungGeneExpr/RData/GCF-2015-373_LungGeneExpr.RData")
## Get an overview of which objects is in the data file:
ls()
## Investigate the lumi-object:
lobj
```

```{r let data get the right format}
## Set the rownames of negativeCtrls to the ProbeID:
rownames(negativeCtrls) <- negativeCtrls$ProbeID

# ## Extract only the LABNR to send to Morten (to obtain additional data):
#
# uniqueLABNR <- sampleNames(lobj)[grep("POOL",sampleNames(lobj),invert=T)]
# ## Save to file:
# write.csv(uniqueLABNR, file="labnr-of-lung-cancer-data.csv",quote=F,row.names=F,col.names=F)

# Convert factors to character (R defaults to changing all txt to factors when reading csv files)
i <- sapply(overview.data, is.factor) #identify factors
overview.data[i] <- lapply(overview.data[i], as.character) #make character using factor levels
i <- sapply(BloodCancer, is.factor) #identify factors
BloodCancer[i] <- lapply(BloodCancer[i], as.character) #make character using factor levels

## Investigate the data:
table(overview.data$Plate)
## Take a look at how the indiviuals with LABNR POOLX are distriubed, 
## and save their information in separate objects:
poolID <- grep("POOL",overview.data$Sample_ID)
poolNegCtrl <- negativeCtrls[,grep("POOL",colnames(negativeCtrls))]
poolLabInfo <- overview.data[poolID,]
## Check that the gene expression data and the laboratory information/questionnaire information are sorted in the same order with respect to LABNR:
all(overview.data$Sample_ID == sampleNames(lobj)) # If TRUE, the OK.
all(overview.data$Sample_ID == BloodCancer$labnr) ## not true. 


## Investigate the data by visual inspection:
overview.data[c(1:5,poolID),]
## Most of the "POOLX" individuals does not seem to have a value for "Plate". But they do have a value for "Chip". Lets look at it:
table(overview.data$Chip) ## Chip value for all women.
table(overview.data[poolID,"Chip"]) #Chip value for pool ID women.

## For now, we remove the POOLX women, and continue to look at the other data:
data <- lobj[,-poolID]
labInfo <- overview.data[-poolID,]
rownames(labInfo) <- labInfo$Sample_ID
negCtrl <- negativeCtrls[,-grep("POOL",colnames(negativeCtrls))]
dim(negCtrl)
## Store the POOLX data in a separate lumi-object:
poolData <- lobj[,poolID]
## Visual inspection of the pool data: 
farge <- rep(2:4,each=4)
boxplot(log2(exprs(poolData)),col=farge)
title("Two pools per chip, and four pools per plate \n log2 of raw data")
## We observe some small batch effects. The pool data should be included in the background correction and normalization procedure.
## Zoomed boxplot:
boxplot(log2(exprs(poolData)),col=farge,ylim=c(5,8))
title("Pools, zoomed")
## Now check if the data in the BloodCancer object is sorted in the same order as the gene expression data: 
all(labInfo$Sample_ID == BloodCancer$labnr) ## Still not true. 
## Sort BloodCancer in the same order as the gene expression data:
tt <- match(labInfo$Sample_ID,BloodCancer$labnr) ## get the order of labnr in Bloodcancer relative to "labInfo"
# Does the "tt" look reasonable:
head(tt)
length(tt)
any(is.na(tt))

## Sort BloodCancer according to the order of "labInfo":
ttt <- BloodCancer[tt,]
ttt[1:10,1:10]
## Check:
all(labInfo$Sample_ID == ttt$labnr) # if true, then ok.
## Check #2 (extract a random individual, and check that the data are similar): 
BloodCancer[BloodCancer$labnr=="100515",]
all(ttt[ttt$labnr=="100515",]== BloodCancer[BloodCancer$labnr=="100515",],na.rm=T)
## Rename: 
background <- ttt
rownames(background) <- background$labnr

## View data:
table(background$Case_ctrl)
case_labnr <- rownames(background)[background$Case_ctrl=="case"] ## only cases.
ctrl_labnr <- as.character(background[case_labnr,"match_labnr"])

table(background[case_labnr,"Case_ctrl"])  
table(background[ctrl_labnr,"Case_ctrl"]) 

## Dates must be set to the right format:
datoer <- c(grep("dato",colnames(background)),grep("date",colnames(background)),grep("DATE",colnames(background)),grep("DATO",colnames(background)))
lapply(background[,datoer],class) ## All are characters
## Visual inspection:
background[1:5,datoer] ## All have the format DDMMMYYYY
## Change format, but save as a temporary variable first:
bb <- background
for(i in datoer){
  bb[,i] <- as.Date(bb[,i], "%d%b%Y")
}
## Visually check that transformation is right by comparing the output:
bb[1:5,datoer]
background[1:5,datoer] 
## ok.

## Update the background data frame:
background[,datoer] <- bb[,datoer]
## Small check:
lapply(background[,datoer],class)

## Find time from blood sample to first diagnosis of lung cancer:
## Subtract time from diagnosis to blood sample, divide by 365.25 and round up to nearest value:
followUpTime <- (background$diagnosedato1 - background$BPROVEDATO)/365.25
followUpY <- ceiling(followUpTime)
table(followUpY)
## Add variables to the "background" data frame:
background$followUpY <- followUpY
background$followUpTime <- followUpTime

## Look at the diagnosis code:
table(background[,"lok_icd71"]) ## just a check to see that controls do not have any diagnosis. 
table(background[case_labnr,"lok_icd71"])

## Look at the metastasis status:
table(background[ctrl_labnr,"metastase1"])## just a check to see that controls do not have any metastasis. 
table(background[case_labnr,"metastase1"])

## Metastasis by follow-up year:
by(background$metastase1,background$followUpY,table)

## In labInfo, all decimal separators are with a comma instead of dot:
head(labInfo)
dd <- grep(",",labInfo[1,])
lapply(labInfo[,dd],class) ## All are characters, but should be numeric.

## one value is with brackets. Remove brackets:
labInfo$RIN[178]
labInfo$RIN[178] <- "6,8"
labInfo$RIN[178]

ll <- labInfo ## Make a temporary object.
## Change comma to dot:
ll[,dd] <- lapply(labInfo[,dd],function(x) as.numeric(gsub(patt=",", replace=".", x)))
labInfo <- ll ## update labInfo.

#############################
#### Save DATA ############
#############################

## Save data to file:
files <- paste0("lung_cancer_raw_data__n=",n,"_p=",p,"_15FEB2016.Rdata")
if(file.exists(files))
{
  print("File already exists. Can not save.")
} else {
  print(paste("saving data to file",files,sep=" "))
  save(data, background, labInfo, negCtrl, poolData, file=files)  
}




```

```{r blood genes}
## For the preprocessing, I remove the blood genes: 
## Remove probes related to blood sample: 
## Import a list (from Arnar) with probes that are related to 
## blood. 
histo <- as.character(read.table("/home/hbo010/breastCancer/data/hla_hist.txt")$V1)
histo
##The values of the list is of from IlluminaID, so we need 
## to convert them to nuID.
histo_nuID <- IlluminaID2nuID(histo, lib.mapping=NULL, species = "Human")[,"nuID"] ## make logical vector of columns to keep:
gExprs<-exprs(data) ## All gene expression
histo_logic <- !rownames(gExprs) %in% histo_nuID
## Check:
sum(histo_logic) - nrow(gExprs)

## extract submatrix:
data.incl.blood <- data 	## lumi-object, temporary renamed.
data <- data[histo_logic,] ## remember that data is of format p x n
dim(data.incl.blood) - dim(data)
```


The data consists of `r dim(lobj)[2]` individuals and `r dim(lobj)[1]` probes, stored in the LumiBatch-object "lobj". We also have available the objects "negativeCtrls" of size `r dim(negativeCtrls)`, and the two objects "overview.data" (size `r dim(overview.data)`) and "samplesheet.data" (size `r dim(samplesheet.data)`) which both contain information from the laboratory. The "samplesheet.data" seems to contain the same information as "overview.data", and seems to be redundant. There is also a data frame named "BloodCancer" of size `r dim(BloodCancer)` containing information from the blood questionnaire as well as from the Cancer Registry.


```{r Round 1 of outlier removal}
##############
### Initial QC.
##############
dataQ <- lumiQ(data)
dataQ
summary(dataQ, "QC")

plotQc <- TRUE 

## Better way to make output plots:
if (plotQc == TRUE) {
	## density plots
  jpeg(filename=file.path(resultPath,"dens.jpeg"), width=2400, height=2400, units="px", quality=85)
  plot(dataQ, what="density"); dev.off()
	## sample relations based on MDS (multidimensional scaling)
  jpeg(filename=file.path(resultPath,"sarel.jpeg"), width=2400, height=2400, units="px", quality=85)
  plot(dataQ, what="sampleRelation"); dev.off()
	## Boxplots
  #jpeg(filename=file.path(resultPath,"box.jpeg"), width=2400, height=2400, units="px", quality=85)
  #plot(dataQ, what="boxplot"); dev.off()
	## detect outlier based on the distance to the mean profile
  jpeg(filename=file.path(resultPath,"tree.jpeg"), width=2400, height=2400, units="px", quality=85)
  plot(dataQ, what="outlier"); dev.off()
}



###################
## PCA plots
###################

## Use the log2 of probe intensities:
exprs<-t(exprs(data)) # Extract the gene expression values, and transpose to obtain format (n x p). This is needed for the prcomp() below.
dim(exprs)
log2.exprs <- log2(exprs)
n <- nrow(exprs) ## number of individuals
p <- ncol(exprs) ## number of probes

## Run pca to investigate the first two principal components:
pcaResult <- prcomp(log2.exprs)
pcData <- data.frame(pcaResult$x)
## look at how much the first few pc-components explain of the variation in the data:
pcVar <- (pcaResult$sdev)^2 / sum(pcaResult$sdev^2) ## percent per component.
## Look at screeplot:
screeplot(pcaResult,type="l") ## Seems that the first two principal componetns are sufficient to consider.

# set variables for printing
y.max<- x.max <-  max(abs(c(max(pcData[1:2]), min(pcData[1:2]))))
y.min <- x.min <- (- y.max)

## PCA plot of the first two directions:
#pdf(file.path(resultPath,paste("pca-plot-n=",n,"-lungCancer.pdf",sep="")))
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point()+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
#dev.off()

## Possible outliers based on PCA plot:
pca.out <- rownames(log2.exprs)[which(abs(pcData[1]) > 75)]
pca.out ## Print LABNR of the outlier(s).

## color pca.out in pca plot:
possible_outlier <- as.factor(ifelse(rownames(log2.exprs) %in% c(pca.out),"outlier?","ok"))
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=possible_outlier))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
#

###########################
## Want to mark the indivdiual(s) above the red line in the tree.jpg.
#	Need to look at plot and decide which ones to investigate - the subjective part of the process
###########################
## There are no obvious outliers in the tree.jpg plot for these data.

############################
## Density plots
############################
## Color the possible outliers with a different color in the density plots.
# Now will make density plot with outliers from above colored red, one plot for each Plate
  color.out.all <- ifelse(rownames(log2.exprs)%in%pca.out,2,1) 
  names(color.out.all) <- rownames(log2.exprs)
  table(color.out.all)
  
  uu <- by(rownames(labInfo), labInfo[,"Plate"], print)
  #pdf(file.path(resultPath,paste("density-plot-per-plate-",nrow(labInfo),"pca-outliers-colored.pdf",sep="")), width=12)
  #par(mfrow=c(3,3))
  for (i in as.integer(names(uu))){
    plot(density(log2.exprs[as.character(uu[[i]][1]),]), main = paste("Plate",i,"n=",length(uu[[i]])),ylim=c(0,4.5),lwd=.3) 
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=color.out.all[j],lwd=.3)
    }
    for(k in pca.out[pca.out %in% uu[[i]]]){ ## We want the colored lines to be printed in front, and to be bolder to make them visible.
      lines(density(log2.exprs[k,]),col=color.out.all[k],lwd=1)
    }
  }
  #dev.off()

  ## Plot all densities in one plot (irrespecitve of plate), and color the pca.outlier to check if it is extreme overall:
  pdf("density-plot-for-all-plates-with-pca-outlier-marked-in-red-zoomed.pdf",width=16)
  plot(dataQ, what="density",col=1,lwd=.3,xlim=c(5.5,7))
  lines(density(log2.exprs[pca.out,]),col="red",lwd=1)
  dev.off()
  ## Looking at the zoomed density plot, It seems that there is one more outlier?! Lets consider the AQM.
  

```

```{r AQM outliers search}
############
####	Another way to look at outliers: AQM identification of outliers.
############

## make data an expressionset to be able to use package arrayQualityMetrics:
eset <- as(data, "ExpressionSet")
class(eset)
preparedData = prepdata(expressionset=eset, intgroup=c(), do.logtransform=TRUE)

# Create object of class aqmReportModule
bo = aqm.boxplot(preparedData)
maplott = aqm.maplot(preparedData, subsample=20000, Dthresh=0.15, maxNumArrays=8, nrColumns=4)
heat = aqm.heatmap(preparedData)
den = aqm.density(preparedData)

# Find the outliers
#	@outliers has 4 slots: statistic, threshold, which (names), description (type of test)
  out.box <- rownames(log2.exprs)[bo@outliers@which]
  out.box
  out.ma <- rownames(log2.exprs)[maplott@outliers@which]
  out.ma
  out.heat <- rownames(log2.exprs)[heat@outliers@which]
  out.heat
  ## no info for density.

  attributes(bo)
  attributes(maplott)
  attributes(heat)
  attributes(den)
  # Find any outliers which are common to all methods  (box=9, ma=39, heat=4)
out.aa <- intersect(out.heat,out.ma);out.aa
out.aqm <- intersect(out.aa,out.box);out.aqm
out.aqm.all <- unique(c(out.heat,out.ma,out.box))

#compare to list found above:
sort(out.aqm)
sort(out.aqm.all)
sort(pca.out)



# ############
# ####	another way to look at outliers: detectOutlier from lumi
# ############
# temp <- detectOutlier(data, ifPlot=FALSE, Th=2) #ifPlot=TRUE makes a plot similar to above, so not needed
# out.dendr <- names(which(temp)) #selects "outliers" based on the lumi definition (no idea)
# sort(out.dendr)

```

```{r visually inspect possible outliers}
#############################
####	Summary of "outliers"
####  ROUND 1
#############################
## Unique outlies found so far: 
out.round1 <- unique(c(out.aqm,pca.out)); out.round1

##################
### PCA PLOT
##################
## Color in PCA plot:
out_score <- rep(1,n)
names(out_score) <- rownames(log2.exprs) ## Individuals not consideres outliers have a value of 1
out_score[pca.out] <- out_score[pca.out]+1 ## outliers from pca have a value of 2
out_score[out.aqm] <- out_score[out.aqm]+2 ## outliers from aqm have a value of 3, and outliers of both pca and aqm have a value of 4.
table(out_score)
## Make factor with understandable names:
out_score <- factor(out_score,levels=1:4,labels=c("ok","PCA outlier","AQM outlier","PCA and AQM"))
table(out_score)

ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=out_score))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))


#####################
### DENSITY PLOT
#####################
## Color in density plot:
 ## Plot all densities in one plot (irrespecitve of plate), and color the pca.outlier to check if it is extreme overall:
  pdf("density-plot-for-all-plates-with-pca-and-aqm-outliers-zoomed.pdf",width=16)
  plot(dataQ, what="density",col=1,lwd=.3,xlim=c(5.5,7))
  for(i in out.round1){  
    lines(density(log2.exprs[i,]),col="red",lwd=1)
  }
dev.off()

############################
## Median probe by outlier:
############################
  ## make median Gene plots for the identified individuals: 
medianGene <- apply(log2.exprs,2,median)
## Set max and min value for plot axis:
minP <- min(log2.exprs)
maxP <- max(log2.exprs)
#pdf(file.path(resultPath,paste("median-geneplot-by-indiv-outliers-r1.pdf",sep="")), width=12)
par(mfrow=c(2,2))
for (i in out.round1){
  plot(log2.exprs[rownames(log2.exprs) %in% i,], medianGene,cex=.3,xlab=paste("GE labnr",i,sep=" "),ylab="",xlim=c(minP,maxP),ylim=c(minP,maxP))
  abline(0, 1) ## diagnoal line showing equality btw mean and individual values.
}
#dev.off()

## For comparison, we will make median gene plots for 8 random individuals in the data:
set.seed(10) ## Set seed (to be able to reproduce the plots later).
ran_ind <- sample(x=rownames(log2.exprs),size = 8,replace = F) ## Select 8 individuals at random.
par(mfrow=c(2,2))
for (i in ran_ind){
  plot(log2.exprs[rownames(log2.exprs) %in% i,], medianGene,cex=.3,xlab=paste("GE labnr",i,sep=" "),ylab="",xlim=c(minP,maxP),ylim=c(minP,maxP))
  abline(0, 1) ## diagnoal line showing equality btw mean and individual values.
}

#############
### LAB INFO
#############
###	Look at Laboratory measures for the individuals of concern
## good quality:	RNA between 50 and 500, RIN>7, 260/280>2, 260/230>1.7 
labInfo[rownames(labInfo) %in% out.round1, c("ng/µl_RNA","RIN","260/280_RNA","260/230_RNA","260/280_cRNA","260/230_cRNA")]
## "140892" has a value below our given threshold for 260/230_RNA.

###	Look at how long in the mail for individuals of concern
days_between <- background[background$labnr %in% out.round1,"freezing_date"]-background[background$labnr %in% out.round1,"BPROVEDATO"]
days_between
## OK.


## PCA plots colored with the different outliers found by different procedures. For comparison:
p1<-  ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.aqm.all))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
p2 <-  ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% pca.out))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max)  + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
p3 <-  ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.dendr))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
p4 <- ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.round1))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
grid.arrange(p1,p2,p3,p4,ncol=2) #this works, just couldn't make it go to a file

```




```{r Check for overall problems (batch effects)}
## Color PCA plot according to batch variables:
## PLATE:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(labInfo$Plate)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
## Case-ctrl status:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(background$Case_ctrl)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## Date RNA isolation:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(labInfo$Date_RNA_Isolation)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## Chip:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(labInfo$Chip)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## RIN:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.numeric(labInfo$RIN)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## Metastase
coll <- ifelse(background$metastase1==0,0,1)
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(coll)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

######################
## DENSITY PLOTS BY PLATE
######################
## Will make three density plots where each plate is the last to be printed in at least one plot. This makes is useful to see that a plate is not very different for the others: 
  uu <- by(rownames(labInfo), labInfo[,"Plate"], print)
  #pdf(file.path(resultPath,paste("density-plot-per-plate-",nrow(labInfo),"pca-outliers-colored.pdf",sep="")), width=12)
  par(mfrow=c(2,2))
  plot(density(log2.exprs[as.character(uu[[1]][1]),]),ylim=c(0,4.5),lwd=.3) 
  for (i in c(1,2,3)){
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=i,lwd=.3)
    }
  }

plot(density(log2.exprs[as.character(uu[[1]][1]),]),ylim=c(0,4.5),lwd=.3) 
  for (i in c(1,3,2)){
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=i,lwd=.3)
    }
  }

plot(density(log2.exprs[as.character(uu[[1]][1]),]),ylim=c(0,4.5),lwd=.3) 
  for (i in c(2,3,1)){
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=i,lwd=.3)
    }
  }

```

```{r Remove outliers}
###	The subjective part here is to decide which, if any, outliers to remove
my.out <- out.round1; my.out

# Datasets: exprs, labInfo, negCtrl, background, data

# getting matches to individuals to remove, so remove the pair
out <- c(my.out, background[background$labnr %in% my.out,"match_labnr"])

## pick out subset of individuals
background<-background[! background$labnr %in% out,]
labInfo<-labInfo[(! labInfo$Sample_ID %in% out),]
negCtrl<-negCtrl[,(! colnames(negCtrl) %in% out)]
data.raw <- data
data<-data[,(! colnames(data) %in% out)] # p x n
controlData(data) <- negCtrl
exprs<-t(exprs(data)) # Extract the gene expression values, and transpose to obtain format (n x p). This is needed for the prcomp() below.


```

```{r Save data}
#############################
#### Save DATA ############
#############################

## Save data to file:
files <- "lung_cancer_after_first_round_outlier_removal.Rdata"
if(file.exists(files))
{
  print("File already exists. Can not save.")
} else {
  print(paste("saving data to file",files,sep=" "))
  save(data, exprs, background, labInfo, negCtrl,out, file=files)  
}


```



```{r Round two of outlier removal}
  
##############
### Initial QC
##############
dataQ <- lumiQ(data)
dataQ
summary(dataQ, "QC")

plotQc <- TRUE 

## Better way to make output plots:
if (plotQc == TRUE) {
	## density plots
  jpeg(filename=file.path(resultPath,"dens2.jpeg"), width=2400, height=2400, units="px", quality=85)
  plot(dataQ, what="density"); dev.off()
	## sample relations based on MDS (multidimensional scaling)
  jpeg(filename=file.path(resultPath,"sarel2.jpeg"), width=2400, height=2400, units="px", quality=85)
  plot(dataQ, what="sampleRelation"); dev.off()
	## Boxplots
  #jpeg(filename=file.path(resultPath,"box.jpeg"), width=2400, height=2400, units="px", quality=85)
  #plot(dataQ, what="boxplot"); dev.off()
	## detect outlier based on the distance to the mean profile
  jpeg(filename=file.path(resultPath,"tree2.jpeg"), width=2400, height=2400, units="px", quality=85)
  plot(dataQ, what="outlier"); dev.off()
}

############
####	another way to look at outliers: detectOutlier from lumi
############
temp <- detectOutlier(data, ifPlot=FALSE, Th=2) #ifPlot=TRUE makes a plot similar to above, so not needed
out.dendr <- names(which(temp)) #selects "outliers" based on the lumi definition (no idea)
sort(out.dendr)



###################
## PCA plots
###################

## Use the log2 of probe intensities:
dim(log2.exprs)

## Run pca to investigate the first two principal components:
pcaResult <- prcomp(log2.exprs)
pcData <- data.frame(pcaResult$x)
## look at how much the first few pc-components explain of the variation in the data:
pcVar <- (pcaResult$sdev)^2 / sum(pcaResult$sdev^2) ## percent per component.
## Look at screeplot:
screeplot(pcaResult,type="l") ## Seems that the first two principal componetns are sufficient to consider.

# set variables for printing
y.max<- x.max <-  max(abs(c(max(pcData[1:2]), min(pcData[1:2]))))
y.min <- x.min <- (- y.max)

## PCA plot of the first two directions:
#pdf(file.path(resultPath,paste("pca-plot-n=",n,"-lungCancer.pdf",sep="")))
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point()+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
#dev.off()

## In this round, I am not sure if I want to call any of the individuals as outliers based on the PCA plot.
## But let's just investigate the ones that have large values on PC1: 
## Possible outliers based on PCA plot:
pca.out <- rownames(log2.exprs)[which(abs(pcData[1]) > 55)]
pca.out ## Print LABNR of the outlier(s).

## color pca.out in pca plot:
possible_outlier <- as.factor(ifelse(rownames(log2.exprs) %in% c(pca.out),"outlier?","ok"))
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=possible_outlier))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))


## Visual output from dendrogram:
out.dendro <- c("106896" ,"103128" ,"128618" ,"135071" ,"146539")
ind.out.dendro <- as.factor(ifelse(rownames(log2.exprs)%in% out.dendro,1,2))
table(ind.out.dendro)
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(col=ind.out.dendro)+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

```

```{r AQM outliers search round 2}
############
####	Another way to look at outliers: AQM identification of outliers.
############

## make data an expressionset to be able to use package arrayQualityMetrics:
eset <- as(data, "ExpressionSet")
class(eset)
preparedData = prepdata(expressionset=eset, intgroup=c(), do.logtransform=TRUE)

# Create object of class aqmReportModule
bo = aqm.boxplot(preparedData)
maplott = aqm.maplot(preparedData, subsample=20000, Dthresh=0.15, maxNumArrays=8, nrColumns=4)
heat = aqm.heatmap(preparedData)
den = aqm.density(preparedData)

# Find the outliers
#	@outliers has 4 slots: statistic, threshold, which (names), description (type of test)
  out.box <- rownames(log2.exprs)[bo@outliers@which]
  out.box
  out.ma <- rownames(log2.exprs)[maplott@outliers@which]
  out.ma
  out.heat <- rownames(log2.exprs)[heat@outliers@which]
  out.heat
  ## no info for density.

  attributes(bo)
  attributes(maplott)
  attributes(heat)
  attributes(den)
  # Find any outliers which are common to all methods  (box=9, ma=39, heat=4)
out.aa <- intersect(out.heat,out.ma);out.aa
out.aqm <- intersect(out.aa,out.box);out.aqm
out.aqm.all <- unique(c(out.heat,out.ma,out.box))

#compare to list found above:
sort(out.aqm)
sort(out.aqm.all)
sort(pca.out)




```

```{r visually inspect possible outliers Round 2}
#############################
####	Summary of "outliers"
####  ROUND 2
#############################
## Unique outlies found so far: 
out.round2 <- unique(c(out.aqm,pca.out,out.dendro)); out.round2

##################
### PCA PLOT
##################
## Color in PCA plot:
## PCA PLOT:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.round2))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## PCA plot for outliers in dendro:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.dendro))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))



## PCA PLOT for all aqm:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.aqm.all))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))


#####################
### DENSITY PLOT
#####################

## Color in density plot:
 ## Plot all densities in one plot (irrespecitve of plate), and color to check if it is extreme overall:
  pdf("density-plot-for-all-plates-with-pca-and-aqm-outliers-zoomed-round2.pdf",width=16)
  plot(dataQ, what="density",col=1,lwd=.3,xlim=c(5.5,7))
  for(i in out.round1){  
    lines(density(log2.exprs[i,]),col="red",lwd=1)
  }
dev.off()

## Color in density plot for out.dendro
 ## Plot all densities in one plot (irrespecitve of plate), and color to check if it is extreme overall:
  pdf("density-plot-for-all-plates-with-dendro-outliers-zoomed-round2.pdf",width=16)
  plot(dataQ, what="density",col=1,lwd=.3,xlim=c(5.5,7))
  for(i in out.dendro){  
    lines(density(log2.exprs[i,]),col="red",lwd=1)
  }
dev.off()

## Color in density plot for out.aqm.all:
 ## Plot all densities in one plot (irrespecitve of plate), and color to check if it is extreme overall:
  pdf("density-plot-for-all-plates-with-aqm-all-outliers-zoomed-round2.pdf",width=16)
  plot(dataQ, what="density",col=1,lwd=.3,xlim=c(5.5,7))
  for(i in out.aqm.all){  
    lines(density(log2.exprs[i,]),col="red",lwd=1)
  }
dev.off()

############################
## Median probe by outlier:
############################
  ## make median Gene plots for the identified individuals: 
medianGene <- apply(log2.exprs,2,median)
## Set max and min value for plot axis:
minP <- min(log2.exprs)
maxP <- max(log2.exprs)
#pdf(file.path(resultPath,paste("median-geneplot-by-indiv-outliers-round2.pdf",sep="")), width=12)
par(mfrow=c(2,2))
for (i in out.round1){
  plot(log2.exprs[rownames(log2.exprs) %in% i,], medianGene,cex=.3,xlab=paste("GE labnr",i,sep=" "),ylab="",xlim=c(minP,maxP),ylim=c(minP,maxP))
  abline(0, 1) ## diagnoal line showing equality btw mean and individual values.
}
#dev.off()

#############
### LAB INFO
#############
###	Look at Laboratory measures for the individuals of concern
## good quality:	RNA between 50 and 500, RIN>7, 260/280>2, 260/230>1.7 
labInfo[rownames(labInfo) %in% out.round2, c("ng/µl_RNA","RIN","260/280_RNA","260/230_RNA","260/280_cRNA","260/230_cRNA")]
## Some individuals have values below our "good quality" limits.

###	Look at how long in the mail for individuals of concern
days_between <- background[background$labnr %in% out.round2,"freezing_date"]-background[background$labnr %in% out.round2,"BPROVEDATO"]
days_between
## OK.

## PCA plots colored with the different outliers found by different procedures. For comparison:
p1<-  ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.aqm.all))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
p2 <-  ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% pca.out))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max)  + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
p3 <-  ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.dendro))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
p4 <- ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=rownames(log2.exprs)%in% out.round2))+ ggtitle(paste("PCA plot of",n,"individuals")) + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
grid.arrange(p1,p2,p3,p4,ncol=2) #this works, just couldn't make it go to a file

```

```{r Check for overall problems (batch effects) Round 2}
## Color PCA plot according to batch variables:
## PLATE:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(labInfo$Plate)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))
## Case-ctrl status:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(background$Case_ctrl)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## Date RNA isolation:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(labInfo$Date_RNA_Isolation)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## Chip:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.factor(labInfo$Chip)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

## RIN:
ggplot(pcData,aes(x=PC1,y=PC2)) + geom_point(aes(color=as.numeric(labInfo$RIN)))+ ggtitle(paste("PCA plot of",n,"individuals"))  + xlim(x.min,x.max) + ylim(y.min,y.max) + xlab(paste("PC1 (",round(pcVar[1]*100)," %)")) + ylab(paste("PC2 (",round(pcVar[2]*100)," %)"))

######################
## DENSITY PLOTS BY PLATE
######################
## Will make three density plots where each plate is the last to be printed in at least one plot. This makes is useful to see that a plate is not very different for the others: 
  uu <- by(rownames(labInfo), labInfo[,"Plate"], print)
  #pdf(file.path(resultPath,paste("density-plot-per-plate-",nrow(labInfo),"pca-outliers-colored.pdf",sep="")), width=12)
  par(mfrow=c(2,2))
  plot(density(log2.exprs[as.character(uu[[1]][1]),]),ylim=c(0,4.5),lwd=.3) 
  for (i in c(1,2,3)){
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=i,lwd=.3)
    }
  }

plot(density(log2.exprs[as.character(uu[[1]][1]),]),ylim=c(0,4.5),lwd=.3) 
  for (i in c(1,3,2)){
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=i,lwd=.3)
    }
  }

plot(density(log2.exprs[as.character(uu[[1]][1]),]),ylim=c(0,4.5),lwd=.3) 
  for (i in c(2,3,1)){
    for (j in as.character(uu[[i]])){
      lines(density(log2.exprs[j,]),col=i,lwd=.3)
    }
  }

```

```{r Remove outliers Round 2}
  ## We choose to not remove any outliers in Round 2. 
  

```

```{r Save data}
#############################
#### Save DATA ############
#############################

## Save data to file:
files <- paste0("lung_cancer_outliers_removed_n=",n,"_p=",p,"_04FEB2016.Rdata")
if(file.exists(files))
{
  print("File already exists. Can not save.")
} else {
  print(paste("saving data to file",files,sep=" "))
  save(data, exprs, background, labInfo, negCtrl, file=files)  
}


```

